---
title: "Permutation Test on Bikeshare Data"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Is there a difference in workday and non-workday daily ridership?

## Dataset

*bike share dataset
from UCI repository: https://archive.ics.uci.edu/ml/datasets/bike+sharing+dataset <br>
Hadi Fanaee-T <br>
Laboratory of Artificial Intelligence and Decision Support (LIAAD), University of Porto <br>
INESC Porto, Campus da FEUP <br>
Rua Dr. Roberto Frias, 378  <br>
4200 - 465 Porto, Portugal* <br>

```{r dataset}
bike<-read.csv("./Data/Bike Sharing - Day.csv")
head(bike)
str(bike)

bk<-bike[,c("workingday","cnt")]
# write.csv(bk,"./Data/bk.csv")

####################################
# bk<-read.csv("./Data/bk.csv")
# head(bk)

#find some summary stats
length(bk$workingday)
sum(bk$workingday)
sum(bk$workingday==0)

hist(bk$cnt,breaks="FD",main = "Histogram of Daily Ridership")
hist(bk$cnt[bk$workingday==1],breaks="FD",main = "Histogram of Workday Ridership")
hist(bk$cnt[bk$workingday==0],breaks="FD",main = "Histogram of Non-Workday Ridership")

tot.mean<-mean(bk$cnt)
tot.med <-median(bk$cnt)
tot.var<-var(bk$cnt)

tot.mean;tot.med;tot.var

wrk.mean<-mean(bk$cnt[bk$workingday==1])
wrk.med <-median(bk$cnt[bk$workingday==1])
wrk.var<-var(bk$cnt[bk$workingday==1])

wrk.mean;wrk.med;wrk.var

off.mean<-mean(bk$cnt[bk$workingday==0])
off.med <- median(bk$cnt[bk$workingday==0])
off.var<-var(bk$cnt[bk$workingday==0])

off.mean;off.med;off.var

obs<-wrk.mean-off.mean;obs
```



## Permutation Test

```{r permutation test}

#permutation test
N<-10000
diffs<-numeric(N)
for (i in 1:N) {
  samp<-sample(bk$workingday)
  wrkAvg<-mean(bk$cnt[samp==1])
  offAvg<-mean(bk$cnt[samp==0])
  diffs[i]<-wrkAvg-offAvg
}
mean(diffs)
hist(diffs, breaks="FD")

abline(v=obs,col="red")
#probability that a difference as large as observed could come from a random subset
pval<-(sum(diffs>=obs)+1)/(N+1) ;pval

```

There is fairly strong evidence that there are more bike share riders on average on workdays. I would reject the null hypothesis that workdays and non workdays have the same ridership at the 95% significance level.




