---
title: "Fit Distributions to Somerville City Data"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# City Wages
Somerville wages: https://data.somervillema.gov/Finance/City-Of-Somerville-Weekly-Payroll-Gross-Wages-Over/q35q-jc9v

```{r Load Data}
SomWage<-read.csv("./Data/City_Of_Somerville_Weekly_Payroll_Gross_Wages_Over_50K_2016.csv")
head(SomWage)
str(SomWage)
SomWage[1,]
#Will use Total Gross. Subtract 50,000 because that is cut off
SomWage.TGC<-SomWage$Total.Gross.Calc-50000

SomWage.hist<-hist(SomWage.TGC,breaks="FD",probability=TRUE,main="Wage Histogram")
```

I will fit gross wages of Somerville city employees to a parametric distribution. The data is truncated at $50000$. It looks like it could follow an exponential distirbution.


```{r Exponential}
m<-mean(SomWage.TGC);m
v<-var(SomWage.TGC);v
lambda<-1/m;lambda
#check variance
((m^2 - v) /v);m^2

plot(SomWage.hist,freq=FALSE, main="Wage Histogram")
curve(dexp(x,lambda), col = "blue", add= TRUE) 
```

The fit looks pretty good, but maybe a Gamma would be better.

```{r Gamma}
beta<-m/v;beta
alpha<-beta*m;alpha

plot(SomWage.hist,freq=FALSE, main="Wage Histogram")
curve(dgamma(x,shape=alpha, rate=beta), col = "forest green", add= TRUE) 
```

The Gamma starts at a low frequency while the dataset starts at a high frequency so the fit is not good. 


Now I use a $\chi^2$-test for goodness of fit.
```{r chi square exp}
#choose bins 
min(SomWage.TGC);max(SomWage.TGC)
plot(SomWage.hist,freq=FALSE, main="Wage Histogram")
abline(v = 6000, col = "red")
abline(v = 15000, col = "red")
abline(v =30000, col = "red")
abline(v =60000, col = "red")
abline(v =100000, col = "red")

#Obs table
ob1<-sum(SomWage.TGC<=6000)
ob2<-sum(SomWage.TGC<=15000)
ob3<-sum(SomWage.TGC<=30000)
ob4<-sum(SomWage.TGC<=60000)
ob5<-sum(SomWage.TGC<=100000)
ob6<-sum(SomWage.TGC>100000)

Observed<-c(ob1,ob2-ob1,ob3-ob2,ob4-ob3,ob5-ob4,ob6);Observed

#expected
expf<-function(x) dexp(x,lambda)
eb1<-integrate(expf,0,6000)$value
eb2<-integrate(expf,6000,15000)$value
eb3<-integrate(expf,15000,30000)$value
eb4<-integrate(expf,30000,60000)$value
eb5<-integrate(expf,60000,100000)$value
eb6<- 1-integrate(expf,0,100000)$value

length(SomWage.TGC)
Expected <- 640*c(eb1,eb2,eb3,eb4,eb5,eb6); Expected

#chi sq
Chi<-sum((Observed-Expected)^2/Expected);Chi #23.30

#how probable get this large of a test stat
#df = 6-1-1=4  6 buckets, calculated lambda from data
Pval<-pchisq(Chi,4,lower.tail=FALSE);Pval #.0001105
#very low
```

I can also perform a similar test but through simulation instead of using a $\chi^2$ distribution directly
```{r simulation Exp}
#draw 10k samples from the exp distribution
N = 10^4; result = numeric(N)
for (i in 1:N) {
  expData = rexp(640,lambda) 
  Counts=numeric(6) 
  Counts[1] = sum(expData <= 6000) 
  Counts[2] = sum((expData > 6000) & (expData <= 15000))
  Counts[3] = sum((expData > 15000) & (expData <= 30000))
  Counts[4] = sum((expData > 30000) & (expData <= 60000))
  Counts[5] = sum((expData > 60000) & (expData <= 100000))
  Counts[6] = sum(expData > 100000)
  result[i] = sum((Counts-Expected)^2/Expected)
}
hist(result, breaks = "FD",probability =TRUE)
curve(dchisq(x, df=4), col = "blue", add= TRUE)
#calculated lambda from the data so should have 4 df but df=5 has a much better fit to simulation
curve(dchisq(x, df=5), col = "Dark Green", add= TRUE) 
abline(v = Chi, col = "red")
sum(result >= Chi)/N;  
```

In both approaches, I get a high test statistic and low probability of getting this distribution from an exponential with this lambda, indicating a poor fit.
<br><br>
# Building Permits
Somerville building permits: https://data.somervillema.gov/City-Services/ISD-Building-Permit-Daily-Applications/q3yh-mp87

I wanted a dataset that could be a good fit for a normal distribution. Each row in this dataset is an application. I'm expecting that the distribution of the number of permit applications each day is approximately normal. 

```{r load data}
Permit<-read.csv("./Data/ISD_Building_Permit_Daily_Applications.csv")
head(Permit)
str(Permit)
min(as.Date(Permit$IssueDate,format = "%m/%d/%Y"))
max(as.Date(Permit$IssueDate,format = "%m/%d/%Y"))
min(as.numeric(as.Date(Permit$IssueDate,format = "%m/%d/%Y")))

IssueDate<-as.numeric(as.Date(Permit$IssueDate,format = "%m/%d/%Y"))-16211
freq<-table(IssueDate)

Permit.hist<-hist(freq,breaks="FD",probability=TRUE,main="Permit Application Histogram")
```

This appears to be more Gamma than Guassian, but it could work as a truncated normal distribution.

```{r fit N}
m<-mean(freq);m
sd<-sd(freq);sd
ln<-length(freq)
plot(Permit.hist, freq=FALSE, main="Permit Application Historgram")
curve(dnorm(x, m, sd), add = TRUE, col = "red")
#actually not bad if we consider a truncated normal distribution

#for truncated normal P(X|X>0)=P(XnX>0)/P(X>0)
fnN<-function(x) dnorm(x,m,sd)
XGT0<-integrate(fnN,0,Inf)$value;trunc #P(X>0)
XLT0<-1-XGT0
m2<-m*XGT0
sd2<-sd*sqrt(XGT0)
curve(dnorm(x, m2, sd2)/XGT0, add = TRUE, col = "blue")
```

Now I do the chi-square test on the truncated normal distribution, bucketing with deciles.

```{r chi sq N}
#10% in each
#.1=P(0<X<a)/P(X>0)
#.1*P(X>0)+P(X<0)=P(X<a)
dec <- qnorm(seq(0.0, 1, by = 0.1)*XGT0+XLT0, m2, sd2);dec
Expected<-rep(ln/10,10); Expected
  
obs<-numeric(10)
for (i in 1:10)
  obs[i] <- sum((freq >= dec[i]) & (freq <= dec[i+1]) ); obs

chi<-sum((obs-Expected)^2/Expected);chi  #16.24
#df=10-1-2=7
curve(dchisq(x, df = 7),from=0,to=50)
abline(v=chi, col = "red")
```

This shows that once again the fitted model is not very good. 


